Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2014-11-24 AO_bondedSi.tex
generalize the use of "puck" to include substrate and sample, closes #4

The reflectivity is about 30\% in the wavelength range 1100 to 5000 nm.  We've defined the coefficient of finesse\cite{2007fuph.book.....S}, $F$, a customary way to encapsulate the Fabry-P\`{e}rot etalon's dependence on the reflectivity.  The coefficient of finesse is shown in the right panel of Figure \ref{figSiIndexFinesse}.  Si is negligibly absorbing longward of about $\lambda$ = 1200 nm, which we verified by comparing the transmission of Si reference samples of different thicknesses, as shown in Figure \ref{figSiAbsorbfig}.  Figure \ref{figSiAbsorbfig} also verifies a key assumption that we will make when comparing the transmission of Si pucks of different heritage and thickness.  The Si refractive index is indistinguishable from sample to sample for the wavelength ranges we care about- the non-absorbing wavelengths greater than about 1200 nm.  \cite{2006SPIE.6273E..77F} cites the wide variety of values for refractive index from the literature as evidence that batch-specific Si should be used as reference for cases in which absolute accuracy greater than $\pm5\times10^{-3}$ is desired.  An absolute deviation of $\pm5\times10^{-3}$ in refractive index corresponds to a Fresnel transmission difference of about 0.2\%, which is comparable to our measurement uncertainty.  In the next subsection we work out the wavelength dependent transmission for the air etalon.

@@

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
There is a key hurdle in computing the transmission spectrum for bonded Si with a finite interface gap- multiple incoherent reflections occur within the Si substrate at normal incidence.  Treatments for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si puck with no gap, and second a pair of bonded Si pucks with a gap thickness $d$.  The DSP Si puck with no air gap has a transmission equal to:

@@

Now consider a pair of bonded Si pucks with a gap thickness $d$. We derive the absolute transmission of bonded Si substrates with a gap in Equation \ref{eqn:Tetalon} in the Appendix.  We normalize the absolute transmission by the transmission with no gap to isolate the effect of the gap.  We call this normalized transmission the etalon transmission $T_{e}$, since it isolates the effect of the etalon on the transmission.

@@

The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify the 3.3 mm thick substrates as ``pucks'' and the $\sim$1.0 mm thick substrates as ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the detector artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

\section{Results of infrared spectroscopy of bonded Si pucks}
\label{secResults}
In this section we describe the spectroscopy of bonded Si pucks.  In some instances IR imaging was available and was leveraged to target or avoid locations demonstrating IR bubbles.  In other cases the IR imaging was not available or ignored, and spectra were taken at random positions on the wafer surface.

@@

\subsection{Spectra of pucks with known gap sizes}
Figures \ref{VS21specm1} to \ref{VS21specf1} show examples of the predicted and measured IR spectra.  Figure \ref{VS21specm1} shows an example spectrum in which the measurement is consistent with the prediction for a gap with the known conditions.  Meanwhile, Figure \ref{VS21specf1} shows a spectrum in which the measurement marginally exceeds the prediction (by about $\sim 2 \sigma$).  Other parts show less consistency with the prediction.  A complete description of the diversity of the measured spectra requires an appeal to optical phenomena beyond the scope of this paper.  Briefly we classify the diversity of measurements into the categories laid out in the table below.

@@

\subsection{Spectra of pucks with no implanted gap}
We have shown that the IR spectroscopy technique combined with the Fabry-P\'{e}rot framework provide accurate knowledge of the gap size.  In the cases where we expected no gap (i.e. perfect Si-Si bond), we have seen spectra consistent with 100\% transmission relative to a DSP Si wafer.  In other cases we have seen evidence for gaps, which is simply an indication that the bonding process introduces gaps.  In a followup paper we will describe ways to mitigate bond formation in bonded Si pucks.

@@

Compare equations \ref{eqn:FPmatTrans} and \ref{eq:FabPerot}.  The revised transmission has picked up a few factors of 2 and $n$.  While we are at it, let's compute the matrix for the scenario with no intermediate etalon: there are simply two Fresnel interfaces with which light interacts incoherently.  This scenario is the model for a single DSP reference puck. The revised matrix multiplication is simply $\boldsymbol{M_{DSP}}=\boldsymbol{M_n}\boldsymbol{M_n}$.  Taking the inverse of the $M_{22}$ element, we find:

>>>>

The reflectivity is about 30\% in the wavelength range 1100 to 5000 nm.  We've defined the coefficient of finesse\cite{2007fuph.book.....S}, $F$, a customary way to encapsulate the Fabry-P\`{e}rot etalon's dependence on the reflectivity.  The coefficient of finesse is shown in the right panel of Figure \ref{figSiIndexFinesse}.  Si is negligibly absorbing longward of about $\lambda$ = 1200 nm, which we verified by comparing the transmission of Si reference samples of different thicknesses, as shown in Figure \ref{figSiAbsorbfig}.  Figure \ref{figSiAbsorbfig} also verifies a key assumption that we will make when comparing the transmission of Si substrates of different heritage and thickness.  Specifically DSP Si pucks and wafers have the indistinguishable transmission longward of 1250 nm.  The Si refractive index is indistinguishable from sample to sample for the wavelength ranges we care about- the non-absorbing wavelengths greater than about 1200 nm.  \cite{2006SPIE.6273E..77F} cites the wide variety of values for refractive index from the literature as evidence that batch-specific Si should be used as reference for cases in which absolute accuracy greater than $\pm5\times10^{-3}$ is desired.  An absolute deviation of $\pm5\times10^{-3}$ in refractive index corresponds to a Fresnel transmission difference of about 0.2\%, which is comparable to our measurement uncertainty.  In the next subsection we work out the wavelength dependent transmission for the air etalon.

@@

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
There is a key hurdle in computing the transmission spectrum for bonded Si with a finite interface gap- multiple incoherent reflections occur within the Si substrate at normal incidence.  Treatments for multiple coherent reflections have been described in detail in the optics literature \cite{2007fuph.book.....S}.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \cite{2002ApOpt..41.3978K}.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si sample with no gap, and second a pair of bonded Si samples with a gap thickness $d$.  The DSP Si sample with no air gap has a transmission equal to:

@@

Now consider a pair of bonded Si samples with a gap thickness $d$. We derive the absolute transmission of bonded Si substrates with a gap in Equation \ref{eqn:Tetalon} in the Appendix.  We normalize the absolute transmission by the transmission with no gap to isolate the effect of the gap.  We call this normalized transmission the etalon transmission $T_{e}$, since it isolates the effect of the etalon on the transmission.

@@

The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify substrates with thicknesses larger than 1 mm as ``pucks''.  Substrates with thickness less than 1.0 mm thick are called ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond pucks with thicknesses as large as 30 mm thick.  These pucks truly resemble over-sized hockey pucks.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the detector artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si samples on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

\section{Results of infrared spectroscopy of directly bonded Si}
\label{secResults}
In this section we describe the spectroscopy of directly bonded Si.  In some instances IR imaging was available and was leveraged to target or avoid locations demonstrating IR bubbles.  In other cases the IR imaging was not available or ignored, and spectra were taken at random positions on the wafer surface.

@@

\subsection{Spectra of direct bonded Si with known gap sizes}
Figures \ref{VS21specm1} to \ref{VS21specf1} show examples of the predicted and measured IR spectra.  Figure \ref{VS21specm1} shows an example spectrum in which the measurement is consistent with the prediction for a gap with the known conditions.  Meanwhile, Figure \ref{VS21specf1} shows a spectrum in which the measurement marginally exceeds the prediction (by about $\sim 2 \sigma$).  Other parts show less consistency with the prediction.  A complete description of the diversity of the measured spectra requires an appeal to optical phenomena beyond the scope of this paper.  Briefly we classify the diversity of measurements into the categories laid out in the table below.

@@

\subsection{Spectra of direct bonded Si with no intentionally implanted gap}
We have shown that the IR spectroscopy technique combined with the Fabry-P\'{e}rot framework provide accurate knowledge of the gap size.  In the cases where we expected no gap (i.e. perfect Si-Si bond), we have seen spectra consistent with 100\% transmission relative to a DSP Si wafer.  In other cases we have seen evidence for gaps, which is simply an indication that the bonding process introduces gaps.

@@

Compare equations \ref{eqn:FPmatTrans} and \ref{eq:FabPerot}.  The revised transmission has picked up a few factors of 2 and $n$.  While we are at it, let's compute the matrix for the scenario with no intermediate etalon: there are simply two Fresnel interfaces with which light interacts incoherently.  This scenario is the model for a single DSP reference sample. The revised matrix multiplication is simply $\boldsymbol{M_{DSP}}=\boldsymbol{M_n}\boldsymbol{M_n}$.  Taking the inverse of the $M_{22}$ element, we find:
